---
title: "Genome 569 Project Notebook"
output: html_document
---

# Analysis Notebook

Use this notebook as your main record for Phase II and Phase III.

This notebook should grow over time. Keep your code, plots, and short notes here.

## Setup

```{r}
library(monocle3)
library(ggplot2)
library(dplyr)
library(readr)
```

## Loading Data

Write a few lines here explaining which data you are using for this phase.

```{r}
counts_dir <- "../provided/full_counts/filtered"
table_s4_path <- "../reference/packer_table_s4.tsv"

# TODO: change counts_dir if your files are stored somewhere else
cds <- load_cellranger_data(counts_dir)
table_s4 <- read_tsv(table_s4_path, comment = "#")

dim(exprs(cds))
head(table_s4)
```

## Quality Control

Start with a simple QC summary of the cells you loaded.

```{r}
colData(cds)$num_genes_expressed <- Matrix::colSums(exprs(cds) > 0)

qplot(colData(cds)$num_genes_expressed, bins = 50) +
  xlab("Genes detected per cell")

# TODO: decide whether any obvious low-quality cells should be removed
```

## Normalization

Describe how you normalized the data.

```{r}
cds <- preprocess_cds(cds, num_dim = 30)

# TODO: change num_dim if needed
```

## PCA and UMAP

Use this section to look at the main structure of the dataset.

```{r}
plot_pc_variance_explained(cds)

cds <- reduce_dimension(cds)
plot_cells(cds, label_groups_by_cluster = FALSE)
```

## Clustering

Identify clusters that may contain the AWA/ASG branch.

```{r}
cds <- cluster_cells(cds)

plot_cells(
  cds,
  color_cells_by = "cluster",
  label_groups_by_cluster = TRUE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
```

## Marker Analysis

Find marker genes for the clusters you think are relevant.

```{r}
marker_genes <- c(
  "odr-7", "nhr-216", "ocr-1",
  "gcy-11", "capa-1",
  "lin-11", "hlh-3", "ces-1", "unc-130", "ztf-11"
)

plot_cells(
  cds,
  genes = marker_genes,
  show_trajectory_graph = FALSE,
  label_cell_groups = FALSE
)

# TODO: decide which clusters best match AWA, ASG, and the shared branch
```

## Identifying the AWA/ASG Branch

Use `reference/packer_table_s4.tsv` to help interpret the clusters.

Useful rows to look for include:

- `ABpxaapap`
- `ABpxaapapa`
- `ABpxaapapaa`

Useful markers to check include:

- AWA: `odr-7`, `nhr-216`, `ocr-1`
- ASG: `gcy-11`, `capa-1`
- shared branch: `lin-11`, `hlh-3`, `ces-1`, `unc-130`, `ztf-11`

```{r}
table_s4 %>%
  filter(Lineage %in% c("ABpxaapap", "ABpxaapapa", "ABpxaapapaa"))

# TODO: match your candidate clusters to these Table S4 rows
# TODO: write a short note explaining your AWA/ASG call
```

## Trajectory Analysis

Now focus on the AWA/ASG branch cells.

```{r}
# TODO: subset branch cells
# TODO: run trajectory or pseudotime analysis
```

## Transcription Factor Analysis

Look for transcription factors that change across the branch.

```{r}
# TODO: start with transcription factors from your marker results
# TODO: compare how they change across the two daughter branches
```

## Final Heatmap

Make a heatmap similar in spirit to Figure 3D from Packer et al., but for AWA vs ASG.

```{r}
# TODO: create the final heatmap
```

## Notes

Write short notes here about:

- how you identified the branch
- which markers were most helpful
- which transcription factors looked most interesting
